---
title: "Using ggpicrust2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using ggpicrust2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

`ggpicrust2` provides a practical workflow for PICRUSt2 downstream analysis:

- convert KO profiles to pathway-level abundance when needed
- run differential abundance analysis with multiple methods
- annotate and visualize pathway-level results
- inspect taxa-level contribution using PICRUSt2 per-sequence outputs
- perform GSEA when pathway-set analysis is more appropriate than single-feature testing

This vignette focuses on the general package workflow. For a deeper GSEA walkthrough, see the dedicated `gsea_analysis` vignette.

## Installation and example data

Install the package and the optional backends used by this tutorial:

```{r installation, eval = FALSE}
install.packages(c("ggpicrust2", "MicrobiomeStat", "BiocManager"))
BiocManager::install(c("KEGGREST", "limma"))
```

```{r example-data, eval = FALSE}
library(ggpicrust2)
library(tibble)
data("ko_abundance")
data("metadata")
alpha <- 0.05
```

Both workflows below use the same data, LinDA method, and adjusted p-value
threshold. LinDA accepts the continuous abundance estimates produced by
`ko2kegg_abundance()`; count-based methods such as ALDEx2 require integer
input and the package rounds non-integer input with a warning. Choose the
method for its assumptions and your study design, not to obtain significance.
KEGG pathway annotation requires internet access and `KEGGREST`.

## One-command workflow

```{r one-command, eval = FALSE}
results <- ggpicrust2(
  data = ko_abundance,
  metadata = metadata,
  group = "Environment",
  pathway = "KO",
  daa_method = "LinDA",
  ko_to_kegg = TRUE,
  order = "pathway_class",
  p_values_bar = TRUE,
  p_values_threshold = alpha,
  x_lab = "pathway_name"
)

# A method's plot is NULL when no pathways can be plotted.
results[[1]]$plot
head(results[[1]]$results)
```

## Stepwise pathway workflow

Run the installation and example-data setup above first. This workflow uses
the same analysis settings as the one-command workflow.

### Convert KO abundance to KEGG pathway abundance

```{r ko-to-kegg, eval = FALSE}
kegg_pathway_abundance <- ko2kegg_abundance(data = ko_abundance)
head(kegg_pathway_abundance[, 1:3])
```

### Match group labels to sample identifiers

`pathway_daa()`, `pathway_heatmap()`, and `pathway_pca()` accept `metadata`
and a column name, such as `group = "Environment"`. In contrast,
`pathway_errorbar()` has no metadata argument: its capitalized `Group`
parameter requires one group label per abundance column, not a column name.

The bundled metadata and abundance table have different sample orders.
Name the group vector with sample IDs so that plotting aligns labels to the
correct samples. For your own data, replace `sample_name` and `Environment`
with your sample-ID and grouping columns. Do not pass an unnamed metadata
column unless you have already verified its order against the abundance columns.

```{r sample-groups, eval = FALSE}
stopifnot(
  !anyNA(metadata$sample_name),
  !anyDuplicated(metadata$sample_name),
  setequal(colnames(kegg_pathway_abundance), metadata$sample_name)
)
sample_groups <- setNames(metadata$Environment, metadata$sample_name)
```

### Run differential abundance analysis

```{r daa, eval = FALSE}
daa_results <- pathway_daa(
  abundance = kegg_pathway_abundance,
  metadata = metadata,
  group = "Environment",
  daa_method = "LinDA"
)

head(daa_results)
```

### Annotate pathway results

```{r annotation, eval = FALSE}
annotated_daa <- pathway_annotation(
  pathway = "KO",
  daa_results_df = daa_results,
  ko_to_kegg = TRUE,
  p_adjust_threshold = alpha
)

head(annotated_daa)
```

### Visualize pathway-level results

`ko_to_kegg = TRUE` is required here too: the rows are KEGG pathways and
use `pathway_name` annotations. This flag does not reconvert the abundance
matrix in the plotting function.

```{r errorbar, eval = FALSE}
sig_pathways <- unique(annotated_daa$feature[
  !is.na(annotated_daa$p_adjust) & annotated_daa$p_adjust < alpha
])

p <- NULL
if (length(sig_pathways) > 0) {
  p <- pathway_errorbar(
    abundance = kegg_pathway_abundance,
    daa_results_df = annotated_daa,
    Group = sample_groups,
    ko_to_kegg = TRUE,
    p_values_threshold = alpha,
    order = "pathway_class",
    x_lab = "pathway_name"
  )
} else {
  message("No pathways pass the adjusted p-value threshold; skipping the error bar plot.")
}
p
```

No significant pathways is a valid analysis outcome. Keep the results table;
do not increase the threshold or change methods just to produce a plot.
Missing KEGG annotations can also prevent plotting even when significant
results exist; check the annotation warnings separately.

```{r heatmap-pca, eval = FALSE}
if (length(sig_pathways) > 0) {
  pathway_heatmap(
    abundance = kegg_pathway_abundance[sig_pathways, , drop = FALSE],
    metadata = metadata,
    group = "Environment"
  )
}

pathway_pca(
  abundance = kegg_pathway_abundance,
  metadata = metadata,
  group = "Environment"
)
```

### Using ALDEx2 instead

ALDEx2 is an optional Bioconductor dependency. For two groups it returns
both Welch and Wilcoxon results, so select one test before annotation and
plotting. Otherwise a pathway can occur twice with different p-values.
Specify the test in advance. ALDEx2 uses Monte Carlo sampling, so set a seed
for reproducibility; results can still differ across package versions.

```{r aldex2-alternative, eval = FALSE}
# Install once with BiocManager::install("ALDEx2").
set.seed(207)
aldex_results <- pathway_daa(
  abundance = kegg_pathway_abundance,
  metadata = metadata,
  group = "Environment",
  daa_method = "ALDEx2"
)
daa_results <- aldex_results[
  aldex_results$method == "ALDEx2_Welch's t test", , drop = FALSE
]
```

Then rerun the annotation and visualization steps with this `daa_results`.
For more than two groups or multiple contrasts, inspect `method`, `group1`,
and `group2` and select the supported test/contrast explicitly. An ALDEx2
run can have no adjusted p-values below 0.05 even when LinDA finds some;
these are different statistical procedures, not equivalent plotting modes.

## Taxa contribution workflow

PICRUSt2 can output contribution files that explain which taxa are driving predicted functional shifts. `ggpicrust2` supports both gene-family-level and pathway-level contribution workflows.

### Read PICRUSt2 contribution files

```{r contrib-readers, eval = FALSE}
# For pred_metagenome_contrib.tsv
contrib_data <- read_contrib_file("pred_metagenome_contrib.tsv")

# For path_abun_contrib.tsv.gz; PICRUSt2 pathway output commonly uses MetaCyc IDs.
path_contrib_data <- read_pathway_contrib_file("path_abun_contrib.tsv.gz")

# For pred_metagenome_strat.tsv
strat_data <- read_strat_file("pred_metagenome_strat.tsv")
```

### Aggregate to a taxonomic level

```{r contrib-aggregate, eval = FALSE}
taxa_contrib <- aggregate_taxa_contributions(
  contrib_data = contrib_data,
  taxonomy = your_taxonomy_table,
  tax_level = "Genus",
  top_n = 10
)

head(taxa_contrib)
```

`aggregate_taxa_contributions()` accepts either:

- normalized contribution data from `read_contrib_file()`
- pathway-level contribution data from `read_pathway_contrib_file()`
- stratified abundance data from `read_strat_file()`

Optional `daa_results_df` or `pathway_ids` filtering requires matching
identifiers and samples. KO-level contribution rows use KO IDs and must not
be filtered with the KEGG pathway-level `daa_results` above; that conversion
does not also convert the contribution file. For MetaCyc contributions, use
DAA results for those same MetaCyc pathways.

When you already have pathway-level contribution data and do not want to run differential abundance analysis, omit `daa_results_df` and aggregate directly. For PICRUSt2 pathway contribution output, use matching MetaCyc annotations rather than KEGG pathway annotations.

```{r pathway-contrib-no-daa, eval = FALSE}
path_taxa_contrib <- aggregate_taxa_contributions(
  contrib_data = path_contrib_data,
  taxonomy = your_taxonomy_table,
  tax_level = "Genus",
  top_n = 10
)

pathway_annotation_df <- pathway_annotation(
  data = data.frame(function_id = unique(path_taxa_contrib$function_id)),
  pathway = "MetaCyc"
)
```

### Visualize taxa-level drivers

```{r contrib-plots, eval = FALSE}
taxa_contribution_bar(
  contrib_agg = taxa_contrib,
  metadata = metadata,
  group = "Environment",
  facet_by = "function"
)

taxa_contribution_heatmap(
  contrib_agg = taxa_contrib,
  n_functions = 20
)
```

This step is useful when pathway-level significance is not enough and you need to identify which taxa are contributing to the change.

## GSEA workflow

Use GSEA when you want pathway-set level inference from KO or EC abundance rather than testing each pathway independently.

```{r gsea, eval = FALSE}
gsea_results <- pathway_gsea(
  abundance = ko_abundance %>% column_to_rownames("#NAME"),
  metadata = metadata,
  group = "Environment",
  pathway_type = "KEGG",
  method = "camera"
)

annotated_gsea <- gsea_pathway_annotation(
  gsea_results = gsea_results,
  pathway_type = "KEGG"
)

visualize_gsea(
  gsea_results = annotated_gsea,
  plot_type = "barplot",
  n_pathways = 15
)
```

For a method-by-method GSEA explanation, covariate adjustment, and comparison with DAA, see the `gsea_analysis` vignette.

## Summary

The package is easiest to use when you choose the shortest path that matches your question:

- use `ggpicrust2()` for a fast default pathway workflow
- use the stepwise DAA functions when you need more control
- use the taxa contribution workflow when you need taxon-level attribution
- use `pathway_gsea()` when pathway-set enrichment is the primary question
